How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-03-07 doc/paper.tex
citation on auc quality. trying to shorten paper

Relative and absolute model adequacy of the four variant models was compared using the area under the receiver operating characteristic curve or AUC \citep{Fawcett2006,Mason2002}. This measure is commonly used in classification problems as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to accuracy which only considers the count of true positives. AUC ranges between 0.5 and 1, with 0.5 indicating no improvement in performance from random and 1 indicating perfect performance. AUC can be interpreted as the probability that our model correctly ranks the relative extinction risks of any two randomly selected species \citep{Fawcett2006,Mason2002}.

The differences in in-sample predictive performance between the models was visualized in multiple ways: whole data set by model, taxonomic group by model, model performance over time, and model performance by taxonomic groups over time. These comparisons demonstrate the relative and absolute adequacy of the models in describing the dataset they were fit to.

@@

The in-sample model comparisons are for determining their relative adequacy, or a model's ability to represent the data it was fit to. Comparison between the posterior predictive estimates of in-sample AUC for each of the four models demonstrates that, overall, all of the models have approximately equal in-sample performance (Fig. \ref{fig:auc_hist}). The parameter rich model VP has the greatest median in-sample AUC when compared to the other three models, but there is substantial overlap in their posterior distributions. Additionally, while our parameter rich model VP is possibly the most adequately performing model, the difference or improvement to performance is minimal at best -- all four models have approximately equal in-sample AUC posterior distributions. All of the in-sample AUC estimates from our models are concentrated around an AUC of 0.77 which is interpreted as ``fine but not good'' performance. It is therefore hard to conclude that there is one ``best'' model which we can rely upon.

@@

Expected out-of-sample predictive performance was estimated using five-fold cross-validation, modified for time series data \citep{ESL}. This procedure yields four posterior (predictive) distributions, each corresponding to AUC values calculated from model-based predictions compared to the extinction state of the hold-out data. These four posterior predictive distributions are pooled to yield a posterior predictive distribution of expected out-of-sample performance -- the resulting distributions tend to be very multimodal due to their very nature being fit to and estimated from different data sets and amounts of data \citep{ESL}. Additionally, multimodality increases with model complexity (Fig. \ref{fig:fold_auc}) -- this makes sense as the more complex models allow for predictor effects to vary with time, allowing for a greater range in possible parameter values which in turn yield a greater range of posterior predictions.

@@

Additionally, the quality of expected out-of-sample performance is not great, with average out-of-sample AUC for each of our models estimated to be between 0.7 and 0.8 which is far from perfect. This result means that we would expect to correctly rank two species in order of most to least likely to go extinct 70-80\% of the time. However, this expected out-of-sample performance is approximately the same as the in-sample performance results (Fig. \ref{fig:auc_hist}), indicating that our models would yield consistent results when generalized to future extinctions.

@@

This result is an important reminder about understanding the practical interpretation of our analyses, which can be lost when we do not consider the predictive aspects of our analyses. By focusing on determining which covariates are ``significant'' and which model is best through simple comparisons means that the practical importance of the results are ignored. For example, in logistic regression a covariate can be considered significant on the log-odds scale but have no practical difference on the probability scale because the range of values is too small as to matter such as when the intercept is greater than 2 or less than -2 because the inverse-logit of those values are close to 0 or 1, respectively \citep{ARM}.

The success of model is partially driven by the size of our dataset and the hierarchical structure of our model. Our estimates are based on rather limited information about the taxa themselves, and our model only takes into account some aspects of species geographic range and their rough taxonomic grouping. Instead of relying on large amounts of ecological information to shape individual differences, our model leverages most of the Neptune dataset through a multilevel model to constrain and improve our parameter estimates by sharing information about those parameters across taxa and time.

>>>>

Relative and absolute model adequacy of the four variant models was compared using the area under the receiver operating characteristic curve or AUC \citep{Fawcett2006,Mason2002}. This measure is commonly used in classification problems as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to accuracy which only considers the count of true positives. AUC ranges between 0.5 and 1, with 0.5 indicating no improvement in performance from random and 1 indicating perfect performance. AUC can be interpreted as the probability that our model correctly ranks the relative extinction risks of any two randomly selected species \citep{Fawcett2006,Mason2002}. AUC values of approximately 0.8 or greater can be considered ``good'' \citep{ACCDA}, which means that values between 0.7 and 0.8 could then be considered ``fair.''

%The differences in in-sample predictive performance between the models was visualized in multiple ways: whole data set by model, taxonomic group by model, model performance over time, and model performance by taxonomic groups over time. These comparisons demonstrate the relative and absolute adequacy of the models in describing the dataset they were fit to.

@@

The in-sample model comparisons are for determining their relative adequacy, or a model's ability to represent the data it was fit to. Comparison between the posterior predictive estimates of in-sample AUC for each of the four models demonstrates that, overall, all of the models have approximately equal in-sample performance (Fig. \ref{fig:auc_hist}). The parameter rich model VP has the greatest median in-sample AUC when compared to the other three models, but there is substantial overlap in their posterior distributions. Additionally, while our parameter rich model VP is possibly the most adequately performing model, the difference or improvement to performance is minimal at best -- all four models have approximately equal in-sample AUC posterior distributions. All of the in-sample AUC estimates from our models are concentrated around an AUC of 0.77. It is therefore hard to conclude that there is one ``best'' model which we can rely upon as they are all nearly functional equivalent.

@@

Expected out-of-sample predictive performance was estimated using five-fold cross-validation \citep{ESL}. This procedure yields four posterior (predictive) distributions, each corresponding to AUC values calculated from model-based predictions compared to the extinction state of the hold-out data. The resulting distribution when all folds are combined is very multimodal by their very nature of being fit to and estimated from different data sets and amounts of data \citep{ESL}. We find the visual degree of this multimodality increases with model complexity (Fig. \ref{fig:fold_auc}), most likely because the complex models allow for predictor effects to vary with time which allows for a greater range in possible parameter values which in turn yield a greater range of posterior predictions.

@@

%This result means that we would expect to correctly rank two species in order of most to least likely to go extinct 70-80\% of the time. However, this expected out-of-sample performance is approximately the same as the in-sample performance results (Fig. \ref{fig:auc_hist}), indicating that our models would yield consistent results when generalized to future extinctions.

@@

%This result is an important reminder about understanding the practical interpretation of our analyses, which can be lost when we do not consider the predictive aspects of our analyses. By focusing on determining which covariates are ``significant'' and which model is best through simple comparisons means that the practical importance of the results are ignored. For example, in logistic regression a covariate can be considered significant on the log-odds scale but have no practical difference on the probability scale because the range of values is too small as to matter such as when the intercept is greater than 2 or less than -2 because the inverse-logit of those values are close to 0 or 1, respectively \citep{ARM}.

The success of our models is partially driven by the size of our dataset and the hierarchical structure of our model. Our estimates are based on rather limited information about the taxa themselves, and our model only takes into account some aspects of species geographic range and their rough taxonomic grouping. Instead of relying on large amounts of ecological information to shape individual differences, our model leverages most of the Neptune dataset through a multilevel model to constrain and improve our parameter estimates by sharing information about those parameters across taxa and time.
